White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-06-19 papers/wdmerger_II/paper.tex
Add notes on problem setup to Paper II

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI}.
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to couple them to ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

@@

Given a set of nuclides and the reaction links between them, we now consider
how a burning step is performed in our software. 
provided with \castro. To \vode\ we provide an integration state containing
the molar abundances $Y_{n} = X_{n} / A_{n}$, where $X$ is the mass fraction
of the abundance in the zone and $A$ is the number of nucleons. The integration
of
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y})
\end{equation}
is determined using a right-hand-side provided by the nuclear burning network.
When it comes to the thermodynamical evolution, the internal energy $e$ of the zone
will change when the nuclear abundances evolve. During the integration we track
the energy release from the system using
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

@@

\subsection{Numerically Unstable Burning}

@@

We implement the white dwarf collision problem in \castro\ using our \wdmerger\
package. The white dwarf centers of mass are separated by a distance of four times
the (secondary) white dwarf radius. Their initial velocity is that of
two point-masses in free-fall towards each other, at the appropriate distance.

@@

Though the main focus of this work is fully three-dimensional simulations,
there is much to be learned from simpler, and much computationally cheaper,
two-dimensional axisymmetric collision simulations. These are implemented
in a cylindrical ($R-z$) coordinate system. We use these for three purposes.
First, in the regime where the 2D and 3D simulations can be usefully compared, 
zero-impact parameter simulations, to understand to what extent the
reduced-dimensionality approach can reproduce the 3D results and thus to
conclude whether 2D simulations can be used for obtaining scientific data
such as nucleosynthesis yields. Second, for testing the effect of various
code choices such as the number of isotopes in the nuclear network. A third
effect which comes for free with the others is that we can benchmark existing
and new code tools such as our nuclear postprocessing analysis. Prepared
with these choices and a better understanding of how reliable our approximations
are, we can then proceed to performing 3D simulations with our code choices
already made and with confidence in our analysis software.

@@

First we briefly consider whether the dynamical evolution of the system
is consistent with simple physics we understand. We ask two questions.
First, is the time it takes for the white dwarfs to collide consistent with
basic kinematics. Second, is the amount of energy released in the
collision by the nuclear reactions comparable to the amount of kinetic
energy we infer from observations of Type Ia supernovae?

>>>>

The hydrodynamical equations solved in this paper were presented in \citet{wdmergerI} (hereafter, Paper I).
In this paper we focus our discussion on our implementation of the nuclear
reaction network and integrator. There are three main areas of concern: first,
specifying what nuclides are in the network and what reaction rates link the
various nuclides; second, once the system of ODEs governing the evolution of
the nuclear species has been written down, how to couple them to ODEs governing
the evolution of the thermodynamics, in particular the temperature; third,
coupling the nuclear reaction updates to the evolution of the hydrodynamical
system. We will deal with each of these in turn in the following subsections.

@@

Given a set of nuclides and the reaction links between them, we now consider
how a burning step is performed in our software. The goal is to integrate the
vector ${\bm{Y}} = (Y_1, Y_2, \ldots, Y_n, e, T)$, where $Y_{n} = X_{n} / A_{n}$
is the molar fraction of species $n$, with $X_n$ the mass fraction and $A_n$ the
mass number of that species, $e$ is the energy released during the burn, and
$T$ is the temperature. The equation describing its evolution is given by
\begin{equation}
  \frac{d\bm{Y}}{dt} = f(\mathbf{Y}),
\end{equation}
where the components of the right-hand-side for the species come from the particular
nuclear burning network we are using. The energy release $e$ of the zone's burn
will change when the nuclear abundances evolve, according to
\begin{equation}
  \frac{\partial e}{\partial t} = N_A \sum_{n} \frac{\partial Y_{n}}{\partial t} m_{n} c^2,
\end{equation}
where $c$ is the speed of light and $m_n$ is the mass of each nuclide.

@@

\subsection{Numerically Unstable Burning}
\label{sec:unstable_burning}

@@

We implement the white dwarf collision problem in \castro\ using our \wdmerger\
package. We load the grid in the way most previous papers have done so:
the white dwarf centers of mass are initially separated by a distance of four times
the (secondary) white dwarf radius. Note that the radius of the white dwarf, and
consequently the initial distance, depends on the equation of state used, and this
can vary somewhat depending on the relevant physics used, such as Coulomb corrections
in the Helmholtz equation of state. However, the results do not really depend on the
exact initial distance as long as there is enough time for the WDs to distort in
response to tidal forces as they approach. Their initial velocity is that of
two point-masses in free-fall towards each other coming in from infinity, such that
the contact point is at the origin, and they approach each other on the $x$-axis.
The composition of the WDs is as described in Paper I: low-mass WDs are helium-dominated,
intermediate-mass WDs are carbon and oxygen dominated, and high-mass WDs contain
oxygen, neon, and magnesium. By default the WDs approach each other head-on, with
their respective centers lying on the $x$-axis, but we have also added an option
to have the WDs approach each other at a non-zero impact parameter $b$, where the
offset is perpendicular to the $x$-axis in the $y$ direction. We measure
$b$ in units of the secondary WD's radius, so that the default head-on case
corresponds to $b = 0$ and the case where the two WDs just graze past each other
corresponds to $b = 1$ (in practice it happens slightly differently because
of the WDs responding to each other's gravitational fields).

For all the 3D simulations to follow, we use the same coarse grid as in Paper I:
there are 256 zones per dimension, and the width of each axis is $1.024 \times 10^{10}$
cm, yielding a base resolution of 400 km. When we use adaptive-mesh refinement, we will
use the refinement scheme described in \autoref{sec:unstable_burning}. This is
acceptable because the outcome of the collision problem is chiefly a function
of the propagation of burning front through the WDs, so for this problem one
can take some liberties in other parts of the algorithm for the purpose of
computational efficiency, which can always be turned back on later if a more
accurate answer is desired. So, for example, we do not need to add refinement
everywhere in the WDs in the interest of getting the collision dynamics to be
slightly more accurate. Additionally, we choose to solve the Poisson equation
for the self-gravity of the system only on the coarse grid. The gravitational
potential and acceleration on the finer levels are interpolated from the coarse
grid.

We have also added to \wdmerger\ the ability to use the 2D cylindrical ($R-z$)
coordinate system evolution in \castro. In this coordinate system, we align
the WDs along the $z$-axis (which is analogous to our $x$-axis in the Cartesian
evolution) at the usual distance, with the center of the WDs at $R = 0$. The
un-simulated $\phi$ dimension then would extend the WDs through a $2\pi$ revolution.
The nature of the axisymmetry inherent to the cylindrical coordinate system
means that we can only run head-on collisions with $b = 0$. The coarse grid
is the same resolution as the 3D case: the width of the $z$ axis is $1.024 \times 10^{10}$
cm and the width of the $R$ axis is $5.012 \times 10^{10}$ cm, with twice as
many zones along the $z$ axis as the $R$ axis so that the equal 400 km
resolution is maintained.

@@

While white dwarf collisions are three-dimensional events, there is much to
be learned from simpler, and much computationally cheaper, two-dimensional
axisymmetric collision simulations. These are implemented in the cylindrical
coordinate system described in \autoref{sec:problemsetup}.
We simulate collisions in 2D cylindrical coordinates for three purposes.
First, we can quickly get up to speed on the nature of the problem and then
benchmark existing and new code tools such as our nuclear postprocessing analysis.
Second, in the regime where the 2D and 3D simulations can be usefully compared, 
zero-impact parameter head-on simulations ($b = 0$), we can understand to
what extent the reduced-dimensionality approach can reproduce the 3D results
and thus determine whether 2D simulations can be used for obtaining scientific
predictions, including nucleosynthesis yields and observable quantities.
Third, we can cheaply test the effect of various code choices such as the number
of isotopes in the nuclear network; we can afford to run the same problem many
times while tweaking parameters one by one to learn their effects. 
Prepared with these results and a better understanding of how reliable our
approximations are, we can then proceed to performing 3D simulations with
our code choices already made and with confidence that we understand the
uncertainty resulting from these choices.

@@

First we briefly consider whether the dynamical evolution of the system
is consistent with simple physics we understand. We ask two questions.
First, is the time it takes for the white dwarfs to collide consistent with
basic kinematics? Second, is the amount of energy released in the
collision by the nuclear reactions comparable to the amount of kinetic
energy we infer from observations of Type Ia supernovae? We must ensure
that we are getting these basics right before we can advance to more
complicated questions. This is not necessarily a trivial matter. We have
observed, for example, that at low resolution, the WDs take longer to
collide than at higher resolution, because the gravitational field is
much less accurate.
